---
title: "Split intuition: matrix examples for two and more populations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Split intuition: matrix examples for two and more populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lantern)
```

## Overview

LANTERN splits variant dosages by local ancestry using one of two methods:

| Mode | Input | Splitting rule |
|------|-------|-------------|
| **Unphased (dosage)** | Genotype matrix + per-variant ancestry code | Probabilistic (p1 … pK proportions estimated per variant) |
| **Phased (haplotype)** | Per-haplotype GT + per-haplotype ancestry | Deterministic — no probability estimation |

This vignette builds intuition for both methods using small, hand-built
matrices — no VCF or MSP files involved. Part A covers the two-population
case; Part B generalizes to K ≥ 3 populations (the same functions, just
with K-population inputs).

* For building the matrices from real VCF + MSP files, see
  `vignette("toy-vcf-msp-example")`.
* For the full split → GDS → association pipeline on real data, see
  `vignette("real-data-chr19")`.

***

# Part A: Two populations

## A.1  Unphased (dosage) split — `split_diploid()`

`split_diploid()` takes two matrices of identical dimensions:

* **`gt_genotype`** — integer matrix, rows = variants, cols = samples.
  Values: 0 (hom-ref), 1 (het), 2 (hom-alt).
* **`ancestry`** — integer matrix, same shape.
  Ancestry codes: **1** = EUR/EUR, **2** = AFR/EUR (mixed het), **3** = AFR/AFR.

The ancestry code summarises *both* haplotypes at the locus (e.g. from an
RFMix MSP file collapsed into a per-variant diploid code).

```{r unphased-basic}
set.seed(42)
n_var  <- 10   # variants
n_samp <- 6    # samples

# Simulate integer genotypes (0/1/2)
gt <- matrix(
  sample(0:2, n_var * n_samp, replace = TRUE, prob = c(0.6, 0.3, 0.1)),
  nrow = n_var, ncol = n_samp
)
rownames(gt) <- paste0("var", seq_len(n_var))
colnames(gt) <- paste0("S", seq_len(n_samp))

# Simulate ancestry codes (1=EUR, 2=MIX, 3=AFR)
anc <- matrix(
  sample(1:3, n_var * n_samp, replace = TRUE, prob = c(0.4, 0.2, 0.4)),
  nrow = n_var, ncol = n_samp
)

result <- split_diploid(gt, anc)

cat("African dosage (first 4 variants × 6 samples):\n")
round(result$african[1:4, ], 2)

cat("\nEuropean dosage (first 4 variants × 6 samples):\n")
round(result$european[1:4, ], 2)
```

For each variant the function computes ancestry-specific proportions
**p1** (African) and **p2** (European) from the diploid counts across all
samples:

```
p1 = (2·N_AA_hom + N_AA_het + N_MIX_hom) / denominator
p2 = (N_MIX_hom + 2·N_EUR_hom + N_EUR_het) / denominator
```

where `denominator = total_alt_alleles − N_MIX_het` (mixed-het samples
are assigned stochastically via p1/p2 rather than excluded from the
denominator).

**Ambiguous-dominated edge-case**: when *all* alt alleles are carried by
mixed-het individuals (the extreme case), `split_diploid()` alone (no
shrinkage target, its default) falls back to p1 = p2 = 0.5 to avoid dividing by zero.
But the more general problem isn't just this all-ambiguous extreme — it's
any variant where mixed-het carriers *dominate* the observations, even with
one or two unambiguous carriers still present to compute a (misleadingly
confident) ratio from. **Part C below explains why that's worth improving
on, and how**; `ancestry_split()` applies that improvement (shrinkage
toward 1/2) automatically.

```{r unphased-edge}
# A variant where only mixed-het (code 2) samples carry the alt allele
gt_edge  <- matrix(c(1L, 0L, 1L), nrow = 1)   # three samples: het, ref, het
anc_edge <- matrix(c(2L, 1L, 2L), nrow = 1)   # all alt carriers are MIX

r <- split_diploid(gt_edge, anc_edge)
cat("Singleton MIX case — p1 = p2 = 0.5\n")
cat("African: ", r$african, "\nEuropean:", r$european, "\n")
```

For matrices with mismatched samples or variant sets,
`ancestry_split_dosage()` wraps `split_diploid()` with automatic overlap
handling and monomorphic filtering — see `?ancestry_split_dosage`.

## A.2  Phased (haplotype) split — `split_haplotype()`

When RFMix (or equivalent) provides **per-haplotype** local ancestry
calls, LANTERN can deterministically assign each haplotype's allele to the
correct ancestry background. Inputs are four matrices of the same shape
(variants × samples):

| Argument | Description |
|----------|-------------|
| `gt_hap0` | integer 0/1 allele for haplotype 0 |
| `gt_hap1` | integer 0/1 allele for haplotype 1 |
| `anc_hap0` | ancestry code for haplotype 0 |
| `anc_hap1` | ancestry code for haplotype 1 |
| `pop_codes` | named vector mapping `c(AFR=0L, EUR=1L)` (RFMix default) |

```{r phased-basic}
# 3 variants × 4 samples
gt0 <- matrix(c(1L, 0L, 1L,   # variant 1: hap0 alleles for 3 samples
                0L, 1L, 0L,   # variant 2
                1L, 1L, 0L),  # variant 3
              nrow = 3, ncol = 3)

gt1 <- matrix(c(0L, 1L, 0L,
                1L, 0L, 1L,
                0L, 1L, 1L),
              nrow = 3, ncol = 3)

# Haplotype ancestry: 0 = AFR, 1 = EUR  (RFMix convention)
a0 <- matrix(c(0L, 0L, 1L,
               1L, 0L, 0L,
               0L, 1L, 0L),
             nrow = 3, ncol = 3)

a1 <- matrix(c(1L, 1L, 0L,
               0L, 1L, 1L,
               1L, 0L, 1L),
             nrow = 3, ncol = 3)

phased_result <- split_haplotype(gt0, gt1, a0, a1)

cat("African dosage (phased):\n"); print(phased_result$african)
cat("European dosage (phased):\n"); print(phased_result$european)
```

The rule is simple: for each cell, hap0 contributes its allele to the AFR
count if `anc_hap0 == 0`, to EUR if `anc_hap0 == 1`, and nothing
otherwise.  Same for hap1 — no probability estimation, just bookkeeping.

***

# Part B: More than two populations

`split_diploid_multi()` and `split_haplotype_multi()` generalize Part A's
functions to K ≥ 2 populations. K = 2 is numerically identical to
`split_diploid()`/`split_haplotype()` — see `?split_diploid_multi` for a
worked equivalence check.

## B.1  Unphased (dosage) split — `split_diploid_multi()`

The diploid ancestry code for K populations encodes which combination of
two parental backgrounds is present.  You supply two objects:

* **`pure_codes`** — named integer vector, one entry per population:
  ```r
  c(AFR = 3L, EUR = 1L, NAT = 4L)
  ```
* **`mixed_codes`** — data frame with columns `code`, `pop1`, `pop2`.
  One row per mixed-ancestry diploid type:
  ```r
  data.frame(
    code = c(2L, 5L, 6L),
    pop1 = c("AFR", "AFR", "EUR"),
    pop2 = c("EUR", "NAT", "NAT")
  )
  ```

For K populations there are K(K−1)/2 mixed-pair types:

| K | Pure types | Mixed pairs | Total codes |
|---|-----------|-------------|-------------|
| 2 | 2 | 1 | 3 |
| 3 | 3 | 3 | 6 |
| 4 | 4 | 6 | 10 |
| 5 | 5 | 10 | 15 |

```{r setup-3pop}
set.seed(2025)
N_VAR  <- 12   # variants
N_SAMP <- 30   # samples

# Ancestry code system
#   Pure:  AFR=3, EUR=1, NAT=4
#   Mixed: AFR/EUR=2, AFR/NAT=5, EUR/NAT=6
pure_codes <- c(AFR = 3L, EUR = 1L, NAT = 4L)
mixed_codes <- data.frame(
  code = c(2L,    5L,    6L   ),
  pop1 = c("AFR", "AFR", "EUR"),
  pop2 = c("EUR", "NAT", "NAT")
)

# Simulate rare-variant genotype matrix (mostly 0, rare 1/2)
gt3 <- matrix(
  sample(0:2, N_VAR * N_SAMP, replace = TRUE, prob = c(0.88, 0.10, 0.02)),
  nrow = N_VAR, ncol = N_SAMP
)
rownames(gt3) <- paste0("chr19:", seq(1e6, by = 500L, length.out = N_VAR))
colnames(gt3) <- paste0("S", sprintf("%02d", seq_len(N_SAMP)))

# Simulate ancestry codes reflecting a tri-admixed cohort
#   Roughly: 30% AFR, 20% EUR, 10% NAT, 20% AFR/EUR, 10% AFR/NAT, 10% EUR/NAT
all_codes <- c(pure_codes, mixed_codes$code)
code_probs <- c(0.30, 0.20, 0.10, 0.20, 0.10, 0.10)
anc3 <- matrix(
  sample(all_codes, N_VAR * N_SAMP, replace = TRUE, prob = code_probs),
  nrow = N_VAR, ncol = N_SAMP
)

out <- split_diploid_multi(gt3, anc3, pure_codes, mixed_codes)

cat("Output list names:", names(out), "\n")
cat("Dimensions (each matrix):", nrow(out$AFR), "variants x",
    ncol(out$AFR), "samples\n\n")

cat("AFR dosage (first 4 variants, 6 samples):\n")
round(out$AFR[1:4, 1:6], 3)

cat("EUR dosage (first 4 variants, 6 samples):\n")
round(out$EUR[1:4, 1:6], 3)

# Conservation check: per-element sum across pools never exceeds raw gt
total <- out$AFR + out$EUR + out$NAT
excess <- max(total - gt3)
cat("\nMax excess of pooled dosage over raw gt:", excess,
    "(must be <= 0)\n")
```

**How p1, p2, p3 are estimated.** For each variant, the function estimates
population allele proportions from **unambiguous observations** only
(pure-ancestry samples of any genotype, plus mixed hom-alts which each
contribute one alt allele to each parent pool).  Ambiguous mixed-ancestry
heterozygotes are excluded from the estimation and then split using the
result.

$$D = \sum_{k} \bigl(2N_{kk}^{(2)} + N_{kk}^{(1)}\bigr) + 2\sum_{i<j} N_{ij}^{(2)}$$

$$p_k = \frac{2N_{kk}^{(2)} + N_{kk}^{(1)} + \sum_{j \neq k} N_{\langle kj \rangle}^{(2)}}{D}, \quad \sum_k p_k = 1$$

A mixed-ancestry heterozygote in pair (i, j) then receives:
$p_i/(p_i+p_j)$ of its dosage in pool i and $p_j/(p_i+p_j)$ in pool j.

```{r proportions-demo}
# Trace p_k for a single variant by hand
var_idx <- 3   # pick variant 3

gt_v   <- gt3[var_idx, ]
anc_v  <- anc3[var_idx, ]

# Pure AFR (code 3)
N_AA2 <- sum(anc_v == 3 & gt_v == 2)
N_AA1 <- sum(anc_v == 3 & gt_v == 1)
# Pure EUR (code 1)
N_EE2 <- sum(anc_v == 1 & gt_v == 2)
N_EE1 <- sum(anc_v == 1 & gt_v == 1)
# Pure NAT (code 4)
N_NN2 <- sum(anc_v == 4 & gt_v == 2)
N_NN1 <- sum(anc_v == 4 & gt_v == 1)
# Mixed hom-alts
N_AE2 <- sum(anc_v == 2 & gt_v == 2)
N_AN2 <- sum(anc_v == 5 & gt_v == 2)
N_EN2 <- sum(anc_v == 6 & gt_v == 2)

D  <- 2*N_AA2 + N_AA1 + 2*N_EE2 + N_EE1 + 2*N_NN2 + N_NN1 +
      2*N_AE2 + 2*N_AN2 + 2*N_EN2
p1 <- if (D > 0) (2*N_AA2 + N_AA1 + N_AE2 + N_AN2) / D else 0.5
p2 <- if (D > 0) (2*N_EE2 + N_EE1 + N_AE2 + N_EN2) / D else 0.5
p3 <- if (D > 0) (2*N_NN2 + N_NN1 + N_AN2 + N_EN2) / D else 0.5

cat(sprintf("Variant %d  D=%d  p_AFR=%.3f  p_EUR=%.3f  p_NAT=%.3f\n",
            var_idx, as.integer(D), p1, p2, p3))
cat("p1+p2+p3 =", p1 + p2 + p3, "\n")
```

**Ambiguous-dominated edge case.** When the denominator D = 0 for a given
variant (all alt alleles are carried exclusively by mixed-ancestry
heterozygotes, with no pure-ancestry or mixed hom-alt alt alleles),
`split_diploid_multi()` alone (no shrinkage target, its default) falls back to
a 50/50 split within each affected pair. That's the extreme case, but the
same unreliability shows up whenever mixed-het carriers dominate the count
even with D > 0 — Part C revisits both.

```{r singleton}
# One sample, one variant: AFR/EUR mixed het, no other data
gt_s  <- matrix(1L, nrow = 1, ncol = 1)
anc_s <- matrix(2L, nrow = 1, ncol = 1)   # code 2 = AFR/EUR
r_s   <- split_diploid_multi(gt_s, anc_s, pure_codes, mixed_codes)
cat("Singleton AFR/EUR het -> AFR:", r_s$AFR[1,1],
    " EUR:", r_s$EUR[1,1],
    " NAT:", r_s$NAT[1,1], "\n")
```

## B.2  Phased (haplotype) split — `split_haplotype_multi()`

When per-haplotype ancestry calls are available (e.g. from RFMix), the
split is **deterministic**: each haplotype's allele goes directly to its
population pool.  No probability estimation is needed. `split_haplotype_multi()`
takes a single named integer vector `pop_codes` of length K, matching the
codes used in the MSP file (RFMix default: 0-based integers).

```{r phased-3pop}
pop_codes_phased <- c(AFR = 0L, EUR = 1L, NAT = 2L)

set.seed(2025)
N_VAR2  <- 8
N_SAMP2 <- 6

# Per-haplotype allele matrices (0/1)
gt_hap0 <- matrix(
  sample(0:1, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.9, 0.1)),
  nrow = N_VAR2, ncol = N_SAMP2
)
gt_hap1 <- matrix(
  sample(0:1, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.9, 0.1)),
  nrow = N_VAR2, ncol = N_SAMP2
)

# Per-haplotype ancestry codes (0=AFR, 1=EUR, 2=NAT)
anc_hap0 <- matrix(
  sample(0:2, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.4, 0.4, 0.2)),
  nrow = N_VAR2, ncol = N_SAMP2
)
anc_hap1 <- matrix(
  sample(0:2, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.4, 0.4, 0.2)),
  nrow = N_VAR2, ncol = N_SAMP2
)

phased_out <- split_haplotype_multi(gt_hap0, gt_hap1,
                                  anc_hap0, anc_hap1,
                                  pop_codes_phased)

cat("Output names:", names(phased_out), "\n")
cat("AFR dosage (first 4 variants):\n")
print(phased_out$AFR[1:4, ])

# Exact reconstruction: AFR + EUR + NAT == hap0 + hap1 everywhere
raw_diploid <- gt_hap0 + gt_hap1
recon_error <- max(abs((phased_out$AFR + phased_out$EUR + phased_out$NAT) -
                         raw_diploid))
cat("Max reconstruction error:", recon_error, "(must be 0)\n")
```

The phased function is exact because every haplotype allele goes to
exactly one pool; there is no probabilistic component.

`split_haplotype_multi()` with K = 2 is numerically identical to
`split_haplotype()`:

```{r phased-2pop-equiv}
pop2 <- c(AFR = 0L, EUR = 1L)
gt0b <- matrix(c(1L,0L,1L,0L), 2, 2)
gt1b <- matrix(c(0L,1L,0L,1L), 2, 2)
a0b  <- matrix(c(0L,1L,0L,1L), 2, 2)
a1b  <- matrix(c(1L,0L,1L,0L), 2, 2)

m2 <- split_haplotype_multi(gt0b, gt1b, a0b, a1b, pop2)
p2 <- split_haplotype(gt0b, gt1b, a0b, a1b, pop2)

cat("AFR identical:", identical(unname(m2$AFR), p2$african), "\n")
cat("EUR identical:", identical(unname(m2$EUR), p2$european), "\n")
```

## B.3  Four- and five-population extension

The same interface extends directly to four or five populations by
adding entries to `pure_codes` and rows to `mixed_codes`. There are
K(K−1)/2 mixed-pair codes for K populations — 6 for K = 4, 10 for K = 5 —
and `mixed_codes` must list all of them.

```{r four-pop}
# Four populations: AFR, EUR, NAT, EAS
pure_4  <- c(AFR = 3L, EUR = 1L, NAT = 4L, EAS = 5L)
mixed_4 <- data.frame(
  code = c(2L, 6L, 7L, 8L, 9L, 10L),
  pop1 = c("AFR","AFR","AFR","EUR","EUR","NAT"),
  pop2 = c("EUR","NAT","EAS","NAT","EAS","EAS")
)

set.seed(11)
n_var4 <- 6; n_samp4 <- 16
gt4  <- matrix(sample(0:2, n_var4 * n_samp4, replace = TRUE,
                       prob = c(0.85, 0.12, 0.03)),
                nrow = n_var4, ncol = n_samp4)
anc4 <- matrix(sample(c(pure_4, mixed_4$code), n_var4 * n_samp4, replace = TRUE),
                nrow = n_var4, ncol = n_samp4)

out4 <- split_diploid_multi(gt4, anc4, pure_4, mixed_4)
cat("Populations:", paste(names(out4), collapse = ", "), "\n")

# Conservation check: pooled dosage never exceeds the raw genotype
total4 <- Reduce("+", out4)
cat("Max excess of pooled dosage over raw gt:", max(total4 - gt4), "(must be <= 0)\n")

# Five populations: AFR, EUR, NAT, EAS, SAS — 10 mixed-pair codes
mixed_5 <- data.frame(
  code = 7:16,
  pop1 = c("AFR","AFR","AFR","AFR","EUR","EUR","EUR","NAT","NAT","EAS"),
  pop2 = c("EUR","NAT","EAS","SAS","NAT","EAS","SAS","EAS","SAS","SAS")
)
cat("K=5 mixed_codes has", nrow(mixed_5), "rows (expected 5*4/2 =", 5 * 4 / 2, ")\n")
```

***

# Part C: Shrinkage toward 1/2 — when ambiguous carriers dominate

Parts A and B each flagged an edge case and deferred it here: the pure
singleton (D = 0, every alt allele carried by mixed-ancestry heterozygotes)
is only the extreme end of a broader problem — the raw formula gets
unreliable well before D actually hits zero, any time ambiguous mixed-het
carriers *dominate* the count. Take 10 AFR/EUR mixed-het carriers (code 2)
and just **one** pure-EUR hom-alt carrier (code 1) at the same variant —
nothing else:

```{r ambiguous-dominant-raw}
gt_dom  <- matrix(c(rep(1L, 10), 2L), nrow = 1)   # 10 mixed hets + 1 pure EUR hom-alt
anc_dom <- matrix(c(rep(2L, 10), 1L), nrow = 1)   # code 2 = AFR/EUR, code 1 = EUR

r_raw <- split_diploid_multi(gt_dom, anc_dom, pure_codes, mixed_codes)  # no shrinkage (gla = NULL)
cat("Raw split -> AFR total:", sum(r_raw$AFR),
    " EUR total:", sum(r_raw$EUR), " NAT total:", sum(r_raw$NAT), "\n")
```

D > 0 here, so this isn't the singleton branch — the raw formula happily
computes p_AFR = 0, p_EUR = 1 from that one unambiguous carrier, and
confidently assigns **all ten** ambiguous carriers' alt alleles to EUR. That's
not a "no answer, guess 50/50" problem; it's a *confidently wrong-looking*
answer extrapolated from a single data point. Nothing in the raw formula
distinguishes "1 unambiguous carrier informing 2 total alt alleles" from "1
unambiguous carrier informing 100" — it takes whatever ratio it computes at
face value, regardless of how little evidence backs it.

**Shrinkage toward 1/2** fixes both this case and the pure singleton with one
mechanism: discount the raw *pairwise* ratio in proportion to how much of the
variant's evidence is actually ambiguous **for that specific pair**, and fill
in the rest with an even split. For the mixed pair `(i,j)` behind a given
ambiguous het (here, AFR/EUR):

$$w_{ij} = \frac{\text{pair } (i,j)\text{'s ambiguous alt alleles}}{\text{all alt alleles at this variant (hom-alt carriers count twice)}}, \qquad
\frac{p_i}{p_i+p_j} \leftarrow (1-w_{ij}) \frac{p_i^{\text{raw}}}{p_i^{\text{raw}}+p_j^{\text{raw}}} + w_{ij} \cdot \frac{1}{2}$$

`w_ij = 0` (pair `(i,j)`'s own unambiguous carriers dominate) reduces to
exactly the raw pairwise ratio from Parts A/B. `w_ij = 1` (the pure singleton
for that pair) gives exactly 1/2 : 1/2 — the old flat fallback, now just one
end of the same continuum rather than a separate rule. Because the blend
stays inside the pair, an ambiguous AFR/EUR het never receives any dosage
toward a third population such as NAT.

**Why 1/2, and not the cohort's ancestry mix?** A mixed-ancestry
heterozygote carries exactly one AFR and one EUR haplotype at the site. Which
of the two holds the allele depends on how common the allele is on AFR versus
EUR haplotypes — not on how many AFR haplotypes the cohort has overall. An
earlier version shrank toward each chromosome arm's "global local ancestry"
(GLA, ~0.82 AFR in the chr19 cohort), which pushed ambiguous alleles toward
AFR. In chr19 simulations the true AFR share of mixed-het alleles was 0.50 at
every allele count, and shrinking toward 1/2 both assigned alleles more
accurately and gave more power when the causal effect was on EUR alleles.

`split_diploid()`/`split_diploid_multi()` accept a shrinkage target through
their optional `gla` (target proportions, one row per group) + `arm_id`
(which row each variant uses) arguments; leaving them at their default
`NULL` reproduces Parts A/B exactly, as seen above. Revisiting the example
with the 1/2 target that `ancestry_split()` uses (1/K for each of K
populations, which is 1/2 for every pair):

```{r half-ambiguous-dominant}
half <- matrix(1/3, nrow = 1, ncol = 3, dimnames = list("all", c("AFR", "EUR", "NAT")))

r_half <- split_diploid_multi(gt_dom, anc_dom, pure_codes, mixed_codes,
                              gla = half, arm_id = 0L)

cat("No shrinkage  -> AFR total:", sum(r_raw$AFR),
    " EUR total:", sum(r_raw$EUR), " NAT total:", sum(r_raw$NAT), "\n")
cat("Toward 1/2    -> AFR total:", round(sum(r_half$AFR), 2),
    " EUR total:", round(sum(r_half$EUR), 2),
    " NAT total:", round(sum(r_half$NAT), 2), "\n")
```

Here `w = 10/12 ≈ 0.83` (10 ambiguous alleles out of 12: the pure-EUR
hom-alt carrier counts twice), so the near-total ambiguity is mostly
discounted: each of the ten mixed-het carriers now gets 5/12 ≈ 0.42 AFR
instead of 0, rather than being handed to EUR outright on the strength of
one sample. The pure singleton from B.1 is the same mechanism at its
extreme (`w = 1`, no unambiguous carriers at all), and lands exactly on the
old flat split:

```{r half-singleton}
gt_s  <- matrix(1L, nrow = 1, ncol = 1)
anc_s <- matrix(2L, nrow = 1, ncol = 1)   # code 2 = AFR/EUR

r_flat   <- split_diploid_multi(gt_s, anc_s, pure_codes, mixed_codes)   # no shrinkage
r_half_s <- split_diploid_multi(gt_s, anc_s, pure_codes, mixed_codes,
                                gla = half, arm_id = 0L)

cat("No shrinkage (flat fallback) -> AFR:", r_flat$AFR[1,1],
    " EUR:", r_flat$EUR[1,1], " NAT:", r_flat$NAT[1,1], "\n")
cat("Toward 1/2 (w = 1)           -> AFR:", r_half_s$AFR[1,1],
    " EUR:", r_half_s$EUR[1,1], " NAT:", r_half_s$NAT[1,1], "\n")
```

In practice you won't build the target by hand — `ancestry_split(mode =
"dosage")` always applies shrinkage toward 1/2 for you:

```r
split <- ancestry_split(vcf_path, msp_path, mode = "dosage")
```

See `?split_diploid_multi` and `?split_haplotype_multi` for full argument
documentation. Next: `vignette("toy-vcf-msp-example")` builds these same
inputs from real VCF + MSP file formats.

## Session information

```{r session}
sessionInfo()
```
