---
title: "Multi-ancestry splitting (K >= 3 populations)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-ancestry splitting (K >= 3 populations)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lantern)
```

## Overview

The standard LANTERN workflow splits variant dosages into two
ancestry-specific pools (African and European).  When the study
cohort has three or more ancestral backgrounds — for example a
Latino cohort with African, European, and Native American
contributions — this two-population assumption is too restrictive.

LANTERN provides two K-population splitting functions:

| Mode | Function | Input | Method |
|------|----------|-------|--------|
| **Unphased** | `split_by_ancestry_multi()` | GT matrix + diploid code matrix | Proportional (p1 … pK estimated per variant) |
| **Phased** | `split_phased_multi()` | Per-haplotype GT + per-haplotype ancestry | Deterministic |

Both return a **named list of K matrices** (variants × samples), one per
population.  The 2-population special case is numerically identical to
`split_by_ancestry()` and `split_phased()` respectively.

---

## 1  Unphased multi-ancestry splitting

### 1.1  Specifying the ancestry code system

The diploid ancestry code for K populations encodes which combination of
two parental backgrounds is present.  You supply two objects:

* **`pure_codes`** — named integer vector, one entry per population:
  ```r
  c(AFR = 3L, EUR = 1L, NAT = 4L)
  ```
* **`mixed_codes`** — data frame with columns `code`, `pop1`, `pop2`.
  One row per mixed-ancestry diploid type:
  ```r
  data.frame(
    code = c(2L, 5L, 6L),
    pop1 = c("AFR", "AFR", "EUR"),
    pop2 = c("EUR", "NAT", "NAT")
  )
  ```

For K populations there are K(K−1)/2 mixed-pair types:

| K | Pure types | Mixed pairs | Total codes |
|---|-----------|-------------|-------------|
| 2 | 2 | 1 | 3 |
| 3 | 3 | 3 | 6 |
| 4 | 4 | 6 | 10 |
| 5 | 5 | 10 | 15 |

### 1.2  Three-population example

```{r setup-3pop}
set.seed(2025)
N_VAR  <- 12   # variants
N_SAMP <- 30   # samples

# Ancestry code system
#   Pure:  AFR=3, EUR=1, NAT=4
#   Mixed: AFR/EUR=2, AFR/NAT=5, EUR/NAT=6
pure_codes <- c(AFR = 3L, EUR = 1L, NAT = 4L)
mixed_codes <- data.frame(
  code = c(2L,    5L,    6L   ),
  pop1 = c("AFR", "AFR", "EUR"),
  pop2 = c("EUR", "NAT", "NAT")
)

# Simulate rare-variant genotype matrix (mostly 0, rare 1/2)
gt <- matrix(
  sample(0:2, N_VAR * N_SAMP, replace = TRUE, prob = c(0.88, 0.10, 0.02)),
  nrow = N_VAR, ncol = N_SAMP
)
rownames(gt) <- paste0("chr19:", seq(1e6, by = 500L, length.out = N_VAR))
colnames(gt) <- paste0("S", sprintf("%02d", seq_len(N_SAMP)))

# Simulate ancestry codes reflecting a tri-admixed cohort
#   Roughly: 30% AFR, 20% EUR, 10% NAT, 20% AFR/EUR, 10% AFR/NAT, 10% EUR/NAT
all_codes <- c(pure_codes, mixed_codes$code)
code_probs <- c(0.30, 0.20, 0.10, 0.20, 0.10, 0.10)
anc <- matrix(
  sample(all_codes, N_VAR * N_SAMP, replace = TRUE, prob = code_probs),
  nrow = N_VAR, ncol = N_SAMP
)
```

```{r split-3pop}
out <- split_by_ancestry_multi(gt, anc, pure_codes, mixed_codes)

cat("Output list names:", names(out), "\n")
cat("Dimensions (each matrix):", nrow(out$AFR), "variants x",
    ncol(out$AFR), "samples\n\n")

cat("AFR dosage (first 4 variants, 6 samples):\n")
round(out$AFR[1:4, 1:6], 3)
```

```{r inspect-sums}
cat("EUR dosage (first 4 variants, 6 samples):\n")
round(out$EUR[1:4, 1:6], 3)

# Conservation check: per-element sum across pools never exceeds raw gt
total <- out$AFR + out$EUR + out$NAT
excess <- max(total - gt)
cat("\nMax excess of pooled dosage over raw gt:", excess,
    "(must be <= 0)\n")
cat("Min dosage value across all pools:", min(out$AFR, out$EUR, out$NAT), "\n")
```

### 1.3  How p1, p2, p3 are estimated

For each variant, the function estimates population allele proportions
from **unambiguous observations** only (pure-ancestry samples of any
genotype, plus mixed hom-alts which each contribute one alt allele to
each parent pool).  Ambiguous mixed-ancestry heterozygotes are excluded
from the estimation and then split using the result.

$$D = \sum_{k} \bigl(2N_{kk}^{(2)} + N_{kk}^{(1)}\bigr) + 2\sum_{i<j} N_{ij}^{(2)}$$

$$p_k = \frac{2N_{kk}^{(2)} + N_{kk}^{(1)} + \sum_{j \neq k} N_{\langle kj \rangle}^{(2)}}{D}, \quad \sum_k p_k = 1$$

A mixed-ancestry heterozygote in pair (i, j) then receives:
$p_i/(p_i+p_j)$ of its dosage in pool i and $p_j/(p_i+p_j)$ in pool j.

```{r proportions-demo}
# Trace p_k for a single variant by hand
var_idx <- 3   # pick variant 3

gt_v   <- gt[var_idx, ]
anc_v  <- anc[var_idx, ]

# Pure AFR (code 3)
N_AA2 <- sum(anc_v == 3 & gt_v == 2)
N_AA1 <- sum(anc_v == 3 & gt_v == 1)
# Pure EUR (code 1)
N_EE2 <- sum(anc_v == 1 & gt_v == 2)
N_EE1 <- sum(anc_v == 1 & gt_v == 1)
# Pure NAT (code 4)
N_NN2 <- sum(anc_v == 4 & gt_v == 2)
N_NN1 <- sum(anc_v == 4 & gt_v == 1)
# Mixed hom-alts
N_AE2 <- sum(anc_v == 2 & gt_v == 2)
N_AN2 <- sum(anc_v == 5 & gt_v == 2)
N_EN2 <- sum(anc_v == 6 & gt_v == 2)

D  <- 2*N_AA2 + N_AA1 + 2*N_EE2 + N_EE1 + 2*N_NN2 + N_NN1 +
      2*N_AE2 + 2*N_AN2 + 2*N_EN2
p1 <- if (D > 0) (2*N_AA2 + N_AA1 + N_AE2 + N_AN2) / D else 0.5
p2 <- if (D > 0) (2*N_EE2 + N_EE1 + N_AE2 + N_EN2) / D else 0.5
p3 <- if (D > 0) (2*N_NN2 + N_NN1 + N_AN2 + N_EN2) / D else 0.5

cat(sprintf("Variant %d  D=%d  p_AFR=%.3f  p_EUR=%.3f  p_NAT=%.3f\n",
            var_idx, as.integer(D), p1, p2, p3))
cat("p1+p2+p3 =", p1 + p2 + p3, "\n")
```

### 1.4  Singleton edge case

When the denominator D = 0 for a given variant (all alt alleles are
carried exclusively by mixed-ancestry heterozygotes, with no
pure-ancestry or mixed hom-alt alt alleles), the function falls back to
a 50/50 split within each affected pair.

```{r singleton}
# One sample, one variant: AFR/EUR mixed het, no other data
gt_s  <- matrix(1L, nrow = 1, ncol = 1)
anc_s <- matrix(2L, nrow = 1, ncol = 1)   # code 2 = AFR/EUR
r_s   <- split_by_ancestry_multi(gt_s, anc_s, pure_codes, mixed_codes)
cat("Singleton AFR/EUR het -> AFR:", r_s$AFR[1,1],
    " EUR:", r_s$EUR[1,1],
    " NAT:", r_s$NAT[1,1], "\n")
```

---

## 2  Phased multi-ancestry splitting

When per-haplotype ancestry calls are available (e.g. from RFMix), the
split is **deterministic**: each haplotype's allele goes directly to its
population pool.  No probability estimation is needed.

### 2.1  Specifying populations

`split_phased_multi()` takes a single named integer vector `pop_codes`
of length K.  The codes must match those in the MSP file
(RFMix default: 0-based integers).

```{r setup-phased}
# RFMix codes: AFR=0, EUR=1, NAT=2
pop_codes_phased <- c(AFR = 0L, EUR = 1L, NAT = 2L)
```

### 2.2  Three-population example

```{r phased-3pop}
set.seed(2025)
N_VAR2  <- 8
N_SAMP2 <- 6

# Per-haplotype allele matrices (0/1)
gt_hap0 <- matrix(
  sample(0:1, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.9, 0.1)),
  nrow = N_VAR2, ncol = N_SAMP2
)
gt_hap1 <- matrix(
  sample(0:1, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.9, 0.1)),
  nrow = N_VAR2, ncol = N_SAMP2
)

# Per-haplotype ancestry codes (0=AFR, 1=EUR, 2=NAT)
anc_hap0 <- matrix(
  sample(0:2, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.4, 0.4, 0.2)),
  nrow = N_VAR2, ncol = N_SAMP2
)
anc_hap1 <- matrix(
  sample(0:2, N_VAR2 * N_SAMP2, replace = TRUE, prob = c(0.4, 0.4, 0.2)),
  nrow = N_VAR2, ncol = N_SAMP2
)

phased_out <- split_phased_multi(gt_hap0, gt_hap1,
                                  anc_hap0, anc_hap1,
                                  pop_codes_phased)

cat("Output names:", names(phased_out), "\n")
cat("AFR dosage (first 4 variants):\n")
print(phased_out$AFR[1:4, ])
```

```{r phased-verify}
# Exact reconstruction: AFR + EUR + NAT == hap0 + hap1 everywhere
raw_diploid <- gt_hap0 + gt_hap1
recon_error <- max(abs((phased_out$AFR + phased_out$EUR + phased_out$NAT) -
                         raw_diploid))
cat("Max reconstruction error:", recon_error, "(must be 0)\n")
```

The phased function is exact because every haplotype allele goes to
exactly one pool; there is no probabilistic component.

### 2.3  Two-population equivalence

`split_phased_multi()` with K = 2 is numerically identical to
`split_phased()`.

```{r phased-2pop-equiv}
pop2 <- c(AFR = 0L, EUR = 1L)
gt0b <- matrix(c(1L,0L,1L,0L), 2, 2)
gt1b <- matrix(c(0L,1L,0L,1L), 2, 2)
a0b  <- matrix(c(0L,1L,0L,1L), 2, 2)
a1b  <- matrix(c(1L,0L,1L,0L), 2, 2)

m2 <- split_phased_multi(gt0b, gt1b, a0b, a1b, pop2)
p2 <- split_phased(gt0b, gt1b, a0b, a1b, pop2)

cat("AFR identical:", identical(unname(m2$AFR), p2$african), "\n")
cat("EUR identical:", identical(unname(m2$EUR), p2$european), "\n")
```

---

## 3  Choosing between unphased and phased multi-ancestry

| | `split_by_ancestry_multi()` | `split_phased_multi()` |
|---|---|---|
| **Ancestry input** | Diploid code per variant | Per-haplotype code from RFMix |
| **Splitting rule** | Proportional (p_k estimated per variant) | Deterministic |
| **Mixed hets** | Split by pairwise ratio p_i / (p_i + p_j) | Not applicable — each haplotype is unambiguous |
| **Reconstruction** | AFR + EUR + ... ≤ raw gt | AFR + EUR + ... = raw gt exactly |
| **Required tools** | None | RFMix (phased VCF + MSP) |
| **Accuracy** | Good for common-to-rare variants | Better for rare variants |

---

## 4  Four- and five-population extension

The same interface extends directly to four or five populations by
adding entries to `pure_codes` and rows to `mixed_codes`.

```{r four-pop, eval=FALSE}
# Four populations: AFR, EUR, NAT, EAS
pure_4  <- c(AFR = 3L, EUR = 1L, NAT = 4L, EAS = 5L)
mixed_4 <- data.frame(
  code = c(2L, 6L, 7L, 8L, 9L, 10L),
  pop1 = c("AFR","AFR","AFR","EUR","EUR","NAT"),
  pop2 = c("EUR","NAT","EAS","NAT","EAS","EAS")
)
# split_by_ancestry_multi(gt, anc, pure_4, mixed_4) → list of 4 matrices

# Five populations: AFR, EUR, NAT, EAS, SAS
pure_5  <- c(AFR=3L, EUR=1L, NAT=4L, EAS=5L, SAS=6L)
# mixed_5: 10 rows (one per pair)
```

There are K(K−1)/2 mixed-pair codes for K populations.  For K = 4 that
is 6 codes; for K = 5 it is 10.  The `mixed_codes` data frame must list
all of them.

See `?split_by_ancestry_multi` and `?split_phased_multi` for full
argument documentation.
